### 2019-12-05 RFCs/HDF5/VOL/user_guide/body.tex
Updated the VOL UG based on feedback.
Former-commit-id: 449ef748a4c26dbcc78a52e5f103b2bcff65f970

Not all public HDF5 API calls pass through the VOL. Only calls which require manipulating storage go through the VOL and require a VOL connector author to implement the appropriate callback. Dataspace, property list, error stack, etc. calls have nothing to do with storage manipulation or querying and do not use the VOL. This may be confusing when it comes to property list calls, since many of those calls set properties for storage. Property lists are just collections of key-value pairs, though, so the VOL is not involved with setting or getting any properties.

@@

An external VOL connector is normally implemented as a dynamically loaded plugin. In order to be used by an application, they must be discoverable and loadable via path settings, etc. These plugins are loaded at runtime via the same HDF5 plugin scheme used for filters. A VOL connector for a file is set in the file access property list (FAPL) used to open it, either via H5Pset\_vol() or a connector-specific API call provided by the connector author.

Many VOL connectors (at various stages of construction) can be found here:

\quad \quad https://bitbucket.hdfgroup.org/projects/HDF5VOL

@@

The native VOL connector is a special case. It is currently implemented as a built-in part of the HDF5 library and cannot be unloaded, nor can its ID be closed.

@@

@@

\subsection{Make Sure The VOL Connector Is In The Search Path}

The default location for all HDF5 plugins is set at configure time when building the HDF5 library. This is true for both CMake and the Autotools.

\subsection{Update Your Code To Load And Use A VOL Connector}

How this is done depends on the VOL connector, which may have special API calls for setup or require appropriate configuration files, but the most generic way to modify a program to use a specific VOL connector would be to {\tt use H5Pset\_vol()} to set the VOL connector in the file access property list (fapl) that will be used to open the file.

You will also need to protect any API calls which are only implemented in the native VOL connector as those calls will fail when using a non-native VOL connector. See the section entitled "Adapting HDF5 Software to Use the VOL", below. A list of native VOL API calls has been included in an appendix.

@@

In place of modifying the source code of your application, you may be able to simply set the {\tt HDF5\_VOL\_CONNECTOR} environment variable (see below). This will automatically use the speicified VOL in place of the native VOL connector.

\section{Loading and Registering Connectors}

Before a VOL connector can be used, it must be discoverable by the library and then loaded and registered.

VOL connector plugins are discovered and loaded by the library using the same mechanism as dataset/group filter plugins (REF). Essentially,

\subsection{registration}

@@

Each of these functions will check to see if an appropriate plugin with a matching name, value, etc. is already loaded and check the plugin path (see above) for matching plugins if this is not true. They return {\tt H5I\_INVALID\_HID} if they are unable to register the connector. Most people will use the {\tt \_by\_name()} or {\tt \_by\_value} versions of this API call. The call that takes an {\tt H5VL\_class\_t} struct is considered a lower-level call this is more commonly used by connector authors. Many VOL connectors will provide a connector-specific init call that will load and register the connector for you.

@@

Note that it's considered an error to close/unload the native VOL connector. The library will prevent this. This means that, for the time being, the native VOL connector will always be available. This may change in the future so that the memory footprint of the native VOL connector goes away when not in use.

@@

There is currently no recommended or library-inherent way to version VOL connectors. A {\tt version} field is included in the {\tt H5VL\_class\_t} struct which holds a connector's information and callbacks, but this has not been utilized in the plugin loading code at this time.

@@

Most connectors will provide a special API call which will set the connector in the fapl. These will often be in the form of {\tt H5Pset\_fapl\_<name>()}. For example, the DAOS VOL (https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/daos-vol) provides a {\tt H5Pset\_fapl\_daos()} API call which will take MPI parameters and make this call. See the connector's documentation or public header file(s) for more information.

@@

@@

\subsection{Specific API Call Substitutions}
\subsection{${\tt H5Fis\_hdf5()} \rightarrow {\tt H5Fis\_accessible()}$}

@@

\subsection{${\tt H5Oget\_info[1|2]()} \rightarrow {\tt H5Oget\_[native|data\_model]\_info()}$}

@@

The {\tt H5Oget\_info1()} and {\tt H5Oget\_info2()} API calls are often used by user code to obtain information about an object in the file, however these calls return a struct which contains native information and are thus currently handled via the native connector's \textit{object optional} callback and so are not handled by most external VOL connectors.

These calls will be separated into two functions: one for getting an object's data model information ({\tt H5Oget\_data\_model\_info()}, which will go through the \textit{object get} callback), and another for returning an object's native file format information ({\tt H5Oget\_native\_info()}, which will go through the native connector's \textit{object optional} callback).

\textbf{This section will be expanded when the RFC for this feature is complete.}

@@

\subsection{List of HDF5 API Calls By Callback}

\begin{longtable}{ |l|>{\raggedright\arraybackslash}l| }
    \hline
    \multicolumn{1}{|c|}{VOL Callback} & \multicolumn{1}{|c|}{HDF5 API Call} \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}FILE} \\
    \hline
    create & H5Fcreate \\
    \hline
    open & H5Fopen \\
    \hline
    get & H5Fget\_access\_plist \\
        & H5Fget\_create\_plist \\
        & H5Fget\_fileno \\
        & H5Fget\_intent \\
        & H5Fget\_name \\
        & H5Fget\_obj\_count \\
        & H5Fget\_obj\_ids \\
    \hline
    specific & H5Fdelete \\
             & H5Fflush \\
             & H5Fis\_accessible \\
             & H5Fis\_hdf5 (deprecated, hard-coded to use native connector) \\
             & H5Fmount \\
             & H5Freopen \\
             & H5Funmount \\
    \hline
    close & H5Fclose \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}GROUP} \\
    \hline
    create & H5Gcreate1 (deprecated) \\
           & H5Gcreate2 \\
           & H5Gcreate\_anon \\
    \hline
    open & H5Gopen1 (deprecated) \\
         & H5Gopen2 \\
    \hline
    get & H5Gget\_create\_plist \\
        & H5Gget\_info \\
        & H5Gget\_info\_by\_idx \\
        & H5Gget\_info\_by\_name \\
        & H5Gget\_num\_objs (deprecated) \\
    \hline
    specific & H5Gflush \\
             & H5Grefresh \\
    \hline
    close & H5Gclose \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}DATASET} \\
    \hline
    create & H5Dcreate1 (deprecated) \\
           & H5Dcreate2 \\
    \hline
    open & H5Dopen1 (deprecated) \\
         & H5Dopen2 \\
    \hline
    read & H5Dread \\
    \hline
    write & H5Dwrite \\
    \hline
    get & H5Dget\_access\_plist \\
        & H5Dget\_create\_plist \\
        & H5Dextend \\
        & H5Dget\_offset \\
        & H5Dget\_space \\
        & H5Dget\_space\_status \\
        & H5Dget\_storage\_size \\
        & H5Dget\_type \\
    \hline
    specific & H5Dextend (deprecated) \\
             & H5Dflush \\
             & H5Drefresh \\
             & H5Dset\_extent \\
    \hline
    close & H5Dclose \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}OBJECT} \\
    \hline
    open & H5Oopen \\
         & H5Oopen\_by\_addr \\
         & H5Oopen\_by\_idx \\
         & H5Oopen\_by\_name \\
    \hline
    copy & H5Ocopy \\
    \hline
    get & N/A \\
    \hline
    specific & H5Odecr\_refcount \\
             & H5Oexists\_by\_name \\
             & H5Oflush \\
             & H5O\_incr\_refcount \\
             & H5Orefresh \\
             & H5Ovisit\_by\_name1 (deprecated) \\
             & H5Ovisit\_by\_name2 \\
             & H5Ovisit1 (deprecated) \\
             & H5Ovisit2 \\
    \hline
    close & H5Oclose \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}LINK} \\
    \hline
    create & H5Glink (deprecated) \\
           & H5Glink2 (deprecated) \\
           & H5Lcreate\_hard \\
           & H5Lcreate\_soft \\
           & H5Lcreate\_ud \\
           & H5Olink \\
    \hline
    copy & H5Lcopy \\
    \hline
    move & H5Gmove (deprecated) \\
         & H5Gmove2 (deprecated) \\
         & H5Lmove \\
    \hline
    get & H5Gget\_linkval (deprecated) \\
        & H5Lget\_info \\
        & H5Lget\_info\_by\_idx \\
        & H5Lget\_name\_by\_idx \\
        & H5Lget\_val \\
        & H5Lget\_val\_by\_idx \\
    \hline
    specific & H5Gunlink (deprecated) \\
             & H5Ldelete \\
             & H5Ldelete\_by\_idx \\
             & H5Lexists \\
             & H5Literate \\
             & H5Literate\_by\_name \\
             & H5Lvisit \\
             & H5Lvisit\_by\_name \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}DATATYPE} \\
    \hline
    commit & H5Tcommit1 (deprecated) \\
           & H5Tcommit2 \\
           & H5Tcommit\_anon \\
    \hline
    open & H5Topen1 (deprecated) \\
         & H5Topen2 \\
    \hline
    get & H5Tget\_create\_plist \\
    \hline
    specific & H5Tflush \\
             & H5Trefresh \\
    \hline
    close & H5Tclose \\
    \hline
    \multicolumn{2}{|>{\raggedright\arraybackslash}p{\linewidth}|}{\cellcolor{gray!35}ATTRIBUTE} \\
    \hline
    create & H5Acreate1 (deprecated) \\
           & H5Acreate2 \\
           & H5Acreate\_by\_name \\
    \hline
    open & H5Aopen \\
         & H5Aopen\_by\_idx \\
         & H5Aopen\_by\_name \\
         & H5Aopen\_idx (deprecated)\\
         & H5Aopen\_name (deprecated) \\
    \hline
    read & H5Aread \\
    \hline
    write & H5Awrite \\
    \hline
    get & H5Aget\_get\_create\_plist \\
        & H5Aget\_info \\
        & H5Aget\_info\_by\_idx \\
        & H5Aget\_info\_by\_name \\
        & H5Aget\_name \\
        & H5Aget\_name\_by\_idx \\
        & H5Aget\_space \\
        & H5Aget\_storage\_size \\
        & H5Aget\_type \\
    \hline
    specific & H5Adelete \\
             & H5Adelete\_by\_idx \\
             & H5Adelete\_by\_name \\
             & H5Aexists \\
             & H5Aexists\_by\_name \\
             & H5Aiterate1 (deprecated) \\
             & H5Aiterate2 \\
             & H5Aiterate\_by\_name \\
             & H5Arename \\
             & H5Arename\_by\_name \\
    \hline
    close & H5Aclose \\
    \hline
\caption{Breakdown of HDF5 API calls by VOL callback}
\end{longtable}

>>>>

Not all public HDF5 API calls pass through the VOL. Only calls which require manipulating storage go through the VOL and require a VOL connector author to implement the appropriate callback. Dataspace, property list, error stack, etc. calls have nothing to do with storage manipulation or querying and do not use the VOL. This may be confusing when it comes to property list calls, since many of those calls set properties for storage. Property lists are just collections of key-value pairs, though, so a particular VOL connector is not required to set or get properties.

@@

A VOL connector can be implemented in several ways: as a shared or static library linked to an application; as a dynamically loaded plugin, implemented as a shared library; and even as an internal connector, built into the HDF5 library itself. This document mostly focuses on external connectors, both libraries and plugins, as those are expected to be much more common than internal implementations.

Many VOL connectors (at various stages of implementation) can be found here:

\quad \quad \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL}

@@

The only current (non-test) internal VOL connector distributed with the library is the native file format connector (the "native VOL connector") which contains the code that handles native HDF5 (*.h5/hdf5) files. In other words, even the canonical HDF5 file format is implemented via the VOL, making it a core part of the HDF5 library and not an optional component which could be disabled. It has not been completely abstracted from the HDF5 library, though, and is treated as a special case. For example, it cannot be unloaded and is always present.

@@

The following steps summarize how one would go about using a VOL connector with an application. More information on particular steps can be found later on in this document.

@@

\subsection{Determine How You Will Set The VOL Connector}

Fundamentally, setting a VOL connector involves modifying the file access propertly list (fapl) that will be used to open or create the file.

There are essentially three ways to do this:

\begin{itemize}
    \item Direct use of {\tt H5Pset\_vol()}
    \item Library-specific API calls that call {\tt H5Pset\_vol()} for you
    \item Use the VOL environment variable, which will also call {\tt H5Pset\_vol()} for you
\end{itemize}

Exactly how you go about setting a VOL connector in a fapl, will depend on the complexity of the VOL connector and how much control you have over the application's source code.

\subsection{If Needed: Update Your Code To Load And Use A VOL Connector}

There are two concerns when modifying the application:

\begin{itemize}
    \item It may be convenient to add connector-specific setup calls to the application.

    \item You will also need to protect any API calls which are only implemented in the native VOL connector as those calls will fail when using a non-native VOL connector. See the section entitled "Adapting HDF5 Software to Use the VOL", below. A list of native VOL API calls has been included in an appendix.
\end{itemize}

In some cases, using the VOL environment variable will work well for setting the connector and any associated storage setup and the application will not use API calls that are not supported by the VOL connector. In this case, no application modification will be necessary.

\subsection{If Using A Plugin: Make Sure The VOL Connector Is In The Search Path}

The default location for all HDF5 plugins is set at configure time when building the HDF5 library. This is true for both CMake and the Autotools. The default locations for the plugins on both Windows and POSIX systems is listed further on in this document.

@@

In place of modifying the source code of your application, you may be able to simply set the {\tt HDF5\_VOL\_CONNECTOR} environment variable (see below). This will automatically use the specified VOL in place of the native VOL connector.

\section{Connector Use}

Before a VOL connector can be set in a fapl, it must be registered with the library ({\tt H5Pset\_vol()} requires the connector's {\tt hid\_t} ID) and, if a plugin, it must be discoverable by the library at run time.

\subsection{Registration}

@@

When used with a plugin, these functions will check to see if an appropriate plugin with a matching name, value, etc. is already loaded and check the plugin path (see above) for matching plugins if this is not true. The functions return {\tt H5I\_INVALID\_HID} if they are unable to register the connector. Most people will use the {\tt \_by\_name()} or {\tt \_by\_value} versions of this API call. The call that takes an {\tt H5VL\_class\_t} struct is considered a lower-level call this is more commonly used by connector authors in connector-specific setup/registration code. Many VOL connectors will provide a connector-specific init call that will load and register the connector for you.

@@

Note that it's considered an error to unload the native VOL connector. The library will prevent this. This means that, for the time being, the native VOL connector will always be available. This may change in the future so that the memory footprint of the native VOL connector goes away when not in use.

@@

There is currently no recommended or library-inherent way to version VOL connectors. A {\tt version} field is included in the {\tt H5VL\_class\_t} struct which holds a connector's information and callbacks, but this is intended for use when versioning the class struct.

@@

Most connectors will provide a special API call which will set the connector in the fapl. These will often be in the form of {\tt H5Pset\_fapl\_<name>()}. For example, the DAOS VOL connector \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/daos-vol} provides a {\tt H5Pset\_fapl\_daos()} API call which will take MPI parameters and make this call. See the connector's documentation or public header file(s) for more information.

@@

Dynamically loaded VOL connector plugins are discovered and loaded by the library using the same mechanism as dataset/group filter plugins. The default locations are:

@@

The VOL was engineered to be as unobtrusive as possible and, when a connector which implements most/all of the data model functionality is in use, most applications should require little, if any, modification. As mentioned in the quick start section, most modifications will probably consist of connector setup code (which can usually be accomplished via the environment variable) and protecting native file format functions.

\subsection{Specific API Call Substitutions}
\subsubsection{${\tt H5Fis\_hdf5()} \rightarrow {\tt H5Fis\_accessible()}$}

@@

\subsubsection{${\tt H5Oget\_info[1|2]()} \rightarrow {\tt H5Oget\_info3()} {\tt H5Oget\_native\_info()}$}

@@

The {\tt H5Oget\_info1()} and {\tt H5Oget\_info2()} family of HDF5 API calls are often used by user code to obtain information about an object in the file, however these calls return a struct which contains native information and are thus currently handled via the native connector's \textit{object optional} callback and so are not handled by most external VOL connectors.

These calls will be separated into two functions: one for getting an object's data model information ({\tt H5Oget\_info3()}, which will go through the \textit{object get} callback), and another for returning an object's native file format information ({\tt H5Oget\_native\_info()}, which will go through the native connector's \textit{object optional} callback). These API calls will use new {\tt H5O\_info2\_t} and {\tt H5O\_native\_info\_t} structs, respectively, to hold the returned data.

This feature is under development and this section will be updated when the feature is complete.

\subsubsection{Optional: ${\tt H5Ovisit[1|2]()} \rightarrow {\tt H5Ovisit3()}$}

The {\tt H5Ovisit()} family of API functions uses the {\tt H5O\_info\_t} struct in its callback function:

\quad \quad {\tt typedef herr\_t (*H5O\_iterate\_t)(hid\_t obj, const char *name, const H5O\_info\_t *info, void *op\_data) }

A new callback that takes the updated, data-model-only struct will replace the old callback in new {\tt H5Ovisit3} API calls. Note that replacing this function is optional as it goes through the VOL's \textit{object specific} callback.

As with the {\tt H5Oget\_info()} work, this feature is under development and this section will be updated when the feature is complete.

@@
